Symbolic Weighted Language Models and Parsing over Infinite Alphabets

### 2021-05-07 main.tex
corresp. swTA - swVPA

%% PGF, Tikz
%% \usepackage{pgfplots}
%% \usepgfplotslibrary{dateplot}
%% \usepackage{pgf,pgfarrows,pgfnodes, pgfautomata}
% \usepackage{tikz}
%% \usetikzlibrary{arrows}
%% \usetikzlibrary{calc}
%% \usetikzlibrary{snakes}
%% \usetikzlibrary{backgrounds}
% \usetikzlibrary{trees}
%% \usetikzlibrary{automata}
%% \usetikzlibrary{positioning}
%% \usetikzlibrary{matrix}
%% \usetikzlibrary{patterns}
%% \usetikzlibrary{shapes}

@@

\def\<#1>{\langle #1 \rangle}
\newcommand{\pair}[2]{\langle{#1}, {#2}\rangle}
%\newcommand{\A}{\mathcal{A}}
%\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\W}{\mathbb{W}}

@@

\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\newcommand{\call}[1]{\ensuremath #1} %{\ensuremath \langle_{#1}}
\newcommand{\return}[1]{\ensuremath #1} %{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\def\weie{{\wei_\mathsf{e}}}
\newcommand{\config}[2]{\ensuremath \genfrac{[}{]}{0pt}{}{#1}{#2}} 
\newcommand{\opluseq}{\ensuremath\mathrel{\oplus}=}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}

@@

@@

In Section~\ref{section:SWA} we introduce \SWA and \SWT, 
then \SWVPA are defined in Section~\ref{sec:SWVPA}, 
where a polynomial 1-best algorithm is described that can be use to solve
Symbolic Weighted Parsing.
Finally, \label{sec:transcription}, we present an application 
of this approach to automated music transcription that has been implemented.

@@

\subsection{Definitions}
%(SWT)
\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over the input and output alphabets~$\Sigma$ and $\Delta$ 
and the semiring $\Semiring$ is a tuple
$T = \< Q, \init, \vec{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\vec{\wei}$ is a tuplet made of the 4 of transition functions 
$\wei_\epsilon$, $\wei_\Sigma$, $\wei_\Delta$, and $\wei_{\Sigma, \Delta}$
from $Q \times Q$ into respectively~$\Semiring$, %$\Phi_\epsilon$, 
$\Phi_\Sigma$, $\Phi_\Delta$, and $\Phi_{\Sigma, \Delta}$.
\end{definition}
%
We synthesize the above 4-uplet of transition functions into a unique 
function~$\wei$ from
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q$
into $\Semiring$, %also called $\wei$ for simplicity, 
such that, for all $q, q' \in Q$, $a \in \Sigma$, and $b \in \Delta$, 
%and with 
%$\< \phi_\epsilon, \phi_\Sigma, \phi_\Delta, \phi_{\Sigma, \Delta}> = \wei(q, q')$, 
\[
\begin{array}{rcll}
\wei(q, \epsilon, \epsilon, q') & = &  \wei_\epsilon(q, q'),\\ %\phi_\epsilon\\
\wei(q, a, \epsilon, q') & = & \phi_\Sigma(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi_\Sigma = \wei_\Sigma(q, q'),\\
\wei(q, \epsilon, b, q') & = & \phi_\Delta(b) & 
\quad\mathrm{where~} \phi_\Delta = \wei_\Delta(q, q'),\\
\wei(q, a, b, q') & = & \phi_{\Sigma, \Delta}(a, b) & 
\quad\mathrm{where~} \phi_{\Sigma, \Delta} = \wei_{\Sigma, \Delta}(q, q').\\
\end{array}      
\]
%More precisely, $Q \times Q$, 
%resp. $Q \times \Sigma \times Q$,
%$Q \times \Delta \times Q$,
%$Q \times \Sigma \times \Delta \times Q$,
%into resp. $\Phi_0$, $\Phi_\Sigma$, $\Phi_\Delta$ $\Phi_{\Sigma, \Delta}$.

The symbolic-weighted transducer $T$ defines a mapping 
from the pairs of strings of $\Sigma^* \times \Delta^*$ 
into the weights of~$\Semiring$,
based on the following intermediate function $\weight_T$
defined recursively for every $q, q' \in Q$, 
%$a \in \Sigma$, $b \in \Delta$ 
for every strings of $s \in \Sigma^*$, $t \in \Delta^*$:
\[
\begin{array}{rccl}
\weight_T(q, s, t, q') & = & 
   \displaystyle\bigoplus_{q'' \in Q} &
   \wei(q, \epsilon, \epsilon, q'') \otimes \weight_T(q'', s, t, q')\\
   & \oplus & \displaystyle\bigoplus_{q'' \in Q} &
              \wei(q'', \epsilon, \epsilon, q') \otimes \weight_T(q, s, t, q'')\\
 & \oplus & \displaystyle\bigoplus_{\begin{array}{c}
                                      \scriptstyle q'' \in Q\\[-2pt]
                                      \scriptstyle s = au, a \in \Sigma
                                      \end{array}} &
   \wei(q, a, \epsilon, q'') \otimes \weight_T(q'', u, t, q')\\
 & \oplus & \displaystyle\bigoplus_{\begin{array}{c}
                                    \scriptstyle q'' \in Q\\[-2pt]
                                    \scriptstyle t = bv, b \in \Delta
                                    \end{array}} &
    \wei(q, \epsilon, b, q'') \otimes \weight_T(q'', s, v, q')\\ 
 & \oplus & \displaystyle\bigoplus_{\begin{array}{c}
                                    \scriptstyle q'' \in Q\\[-2pt]
                                    \scriptstyle s = au, a \in \Sigma\\[-2pt]
                                    \scriptstyle t = bv, b \in \Delta
                                    \end{array}} &
   \wei(q, a, b, q'') \otimes \weight_T(q'', u, v, q').\\ 
\end{array}
\]
%** The application of $\oplus$ can be seen as a ND choice between transitions **
We recall that, by convention, an empty sum with $\oplus$ is $\zero$. 
%
Since $\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei(q, a, b, q'')$ equal to $\zero$ in the above expression 
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and writing $b$ 
(including the case $a = \epsilon$ and $a = \epsilon$).
%These functions $\phi$ act as guards for the transducer's transitions, 
%preventing a transition when they return the absorbing $\zero$ of $\Semiring$.
This is analogous to a transition's guard not satisfied by $\<a, b>$ in 
the case of symbolic transducers.

@@

\noindent
The \emph{Symbolic Weighted Automata} %$A = \< Q, \init, \weight, \final >$
over $\Sigma$ and $\Semiring$ 
are defined similarly as the transducers of Definition~\ref{def:SWT}, 
by simply omitting the output symbols.
%
\begin{definition}
A \emph{symbolic-weighted automaton} (\SWA)
over the input alphabet~$\Sigma$
and the semiring $\Semiring$ is a tuple
$A = \< Q, \init, \vec{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\vec{\wei}$ is a tuplet made of the 2 of transition functions 
$\wei_\epsilon$ and $\wei_\Sigma$ from $Q \times Q$ into 
respectively $\Semiring$ %$\Phi_\epsilon$ 
and $\Phi_\Sigma$.
\end{definition}

@@

\begin{proposition}
Given a \SWT $T$ 
over alphabet $\Sigma$, $\Delta$ and a commutative semiring~$\Semiring$, 
and $s \in \Sigma^*$, 
there exists an effectively constructible \SWA 
$A_{T, s}$ over $\Delta$ and $\Semiring$ 
such that for all $t \in \Delta^*$, $A_{T, s}(t) = T(s, t)$.
\end{proposition}
%
\begin{proof}
Let $T = \< Q, \init, \vec{\wei}, \final >$ 
where $\vec{\wei} \< \wei_\epsilon, \wei_\Sigma, \wei_\Delta, \wei_{\Sigma, \Delta}>$
has an associated synthesized form $\wei$, 
and let $s = s_1 \ldots s_n$ with $s_1, \ldots, s_n \in \Sigma$.
%$n = |s|$ be the length of $s$.
%
Let $Q' = [1..n] \times Q$ be the state set of $A_{T, s}$,
and let, for all $q \in Q$, 
$\init'(1, q) = \init(q)$ and $\init'(i, q) = \zero$ for all $1 < i \leq n$, 
and $\final'(n, q) = \final(q)$ and $\final'(i, q) = \zero$ for all $1 \leq i <  n$.

\noindent
We define the functions $\wei'_\epsilon$ and $\wei'_\Sigma$ for $A_{T, s}$
in their synthesized form: for all $1 \leq i, j \leq n$, $q, q' \in Q$ and $a \in \Sigma$, 
and $b \in \Delta$, 
\[
\begin{array}{lclcl}
\wei'(\< i, q>, \epsilon, \< i, q'>)   & = & \wei(q, \epsilon, \epsilon, q'), & \quad & \\ 
\wei'(\< i, q>, \epsilon, \< i+1, q'>) & = & \wei(q, s_i, \epsilon, q') & & \mathrm{if~} i < n,\\ 
\wei'(\< i, q>, \epsilon, \< j, q'>)   & = & \zero & & \mathrm{if~} j \neq i, j \neq i+1,\\ 
\wei'(\< i, q>, b, \< i, q'>)          & = & \wei(q, \epsilon, b, q'),\\ 
\wei'(\< i, q>, b, \< i+1, q'>)        & = & \wei(q, s_i, b, q') & & \mathrm{if~} i < n,\\ 
\wei'(\< i, q>, b, \< j, q'>)          & = & \zero & & \mathrm{if~} j \neq i, j \neq i+1. 
\end{array}      
\]
Note that the functions defined above belong to the set $\Phi_\Delta$ 
of the underlying label theory, 
by the hypothesis of closure under partial application.
%
The $\SWA$ wanted is $A_{T, s} = \< Q', \init', \vec{\wei'}, \final' >$.
\qed
\end{proof}

@@

\subsection{Definition}
Let $\Omega$ be a countable alphabet 
%finite (large) or infinite,
that we assume partitioned into three subsets 
$\Omegai$, $\Omegac$, $\Omegac$,
respectively called of \emph{internal}, \emph{call} and \emph{return}
symbols.
% \begin{itemize}
% \item a set $\Omegai$ of \emph{internal symbols} denoted $a$,
% \item a set $\Omegac$ of \emph{call symbols} denoted $\call{a}$,
% \item a set $\Omegar$ of \emph{return symbols} denoted $\return{a}$.
% \end{itemize}
Let~$\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative semiring and  
let  $(\Phi_\epsilon, \Phic, \Phir, \Phicr)$ be a label theory over $\Semiring$
%In order to simplify notations, %and following the definition of Section~\ref{section:transducer}, 
%we shall write respectively 
where $\Phic$, $\Phir$ and~$\Phicr$ stand respectively 
for~$\Phi_\Omegac$, $\Phi_\Omegar$ and~$\Phi_{\Omegac, \Omegar}$.
%
Moreover, we extend this theory with a set $\Phii$ 
of unary functions in $\Omegai \to \Semiring$,
closed under $\oplus$ and $\otimes$.

@@

\noindent
The intuition is the following for the above transitions.
\begin{description}
\item $\weii$ : read the input internal symbol $a$, change state to $q'$.
\item $\weic$ : read the input call symbol $\call{c}$, push it to the stack along with $p$, change state to $q'$.
\item $\weir$ : when the stack is not empty, 
      read and pop from stack a pair made of $\call{c}$ and $p$, 
      read the input return symbol $\return{r}$, change state to $q'$.
      In this case, the weight function $\phi_\mathsf{r}$ 
      computes a value of matching between the call and return symbols.
      This value might be $\zero$ in order to express that the symbols do not match.
\item $\weie$ : when the stack is empty, 
      read the input symbol $\call{r}$, change state to $q'$.
\end{description}

We give now a formal definition of these transitions of the automaton $A$
in term of a weight value
computed by an intermediate function $\weight_A$, like in Section~\ref{sec:SWT}.
In the case of a pushdown automaton, a configuration is composed 
of a state $q \in Q$ and a stack content $\gamma \in \Gamma^*$, where $\Gamma = \Omegac \times P$.
Therefore, $\weight_A$ is a function from 
$Q \times \Gamma^* \times \Omega^* \times Q \times \Gamma^*$ into $\Semiring$.
\[
\begin{array}{rcl}
\weight_A\bigl(\config{q}{\gamma}, a\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weii(q, a, q'') 
 \otimes \weight_A\bigl(\config{q''}{\gamma}, u, \config{q'}{\gamma'}\bigr)\\
%
\weight_A\bigl(\config{q}{\gamma}, {\call{c}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-2pt]
                          \scriptstyle p \in P
                          \end{array}}}
 \weic\bigl(q, {\call{c}}, q'', p\bigr) 
 \otimes \weight_A\bigl(\config{q''}{{\call{c}}\, p\cdot \gamma}, u, \config{q'}{\gamma'}\bigr)\\[1mm]
%
\weight_A\bigl(\config{q}{{\call{c}}\, p\cdot \gamma}, {\return{r}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} 
 \weir\bigl(q, {\call{c}}, p, {\return{r}}, q''\bigr) 
 \otimes \weight_A\bigl(\config{q''}{\gamma}, u, \config{q'}{\gamma'}\bigr)\\
%
\weight_A\bigl(\config{q}{\bot}, {\return{r}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weie(q, {\return{r}}, q'') 
 \otimes \weight_A\bigl(\config{q''}{\bot}, u, \config{q'}{\gamma'}\bigr)\\
\end{array}
\]
where $\bot$ denotes the empty stack and ${\call{c}}\, p\cdot \gamma$ 
denotes a stack where the pair made of ${\call{c}} \in \Omegac$ and $p \in P$ is the top symbol 
and $\gamma$ is the rest of stack.

@@

>>>>

%% PGF, Tikz
\usepackage{tikz-cd}
%% \usepackage{pgfplots}
%% \usepgfplotslibrary{dateplot}
%% \usepackage{pgf,pgfarrows,pgfnodes, pgfautomata}
% \usepackage{tikz}
% \usetikzlibrary{cd}
%% \usetikzlibrary{arrows}
%% \usetikzlibrary{calc}
%% \usetikzlibrary{snakes}
%% \usetikzlibrary{backgrounds}
% \usetikzlibrary{trees}
%% \usetikzlibrary{automata}
%% \usetikzlibrary{positioning}
%% \usetikzlibrary{matrix}
%% \usetikzlibrary{patterns}
%% \usetikzlibrary{shapes}

@@

\def\<#1>{\langle #1 \rangle}
\newcommand{\pair}[2]{\langle{#1}, {#2}\rangle}
%\newcommand{\A}{\mathcal{A}}
%\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\W}{\mathbb{W}}
\newcommand{\T}{\mathcal{T}}

@@

\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWTA{\textsf{swTA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\newcommand{\call}[1]{\ensuremath #1} %{\ensuremath \langle_{#1}}
\newcommand{\return}[1]{\ensuremath #1} %{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\def\weie{{\wei_\mathsf{e}}}
\newcommand{\config}[2]{\ensuremath \genfrac{[}{]}{0pt}{}{#1}{#2}} 
\newcommand{\opluseq}{\ensuremath\mathrel{\oplus}=}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}
\newcommand{\rank}{\mathsf{rk}}
\newcommand{\lin}{\mathsf{lin}}

@@

\begin{figure}
\centering
\begin{tikzpicture}
\node (NFA) at (0,4) {%
  \(
  \begin{array}{c} 
  \mathsf{NFA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{B}\\
  q \xrightarrow{a} q' \quad  a \in \Sigma_\mathsf{fin}
  \end{array} 
  \)
};
%
\node (WA) at (-2,2) {%
  \(
  \begin{array}{c} 
  \mathsf{WA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{S}\\
  q \xrightarrow{a, w} q' \quad  a \in \Sigma_\mathsf{fin}, w \in \mathbb{S}
  \end{array} 
  \)
};
%
\node (SA) at (2,2) {%
  \(
  \begin{array}{c} 
  \mathsf{SA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{B}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{B}
  \end{array} 
  \)
};
%
\node (SWA) at (0,0) {%
  \(
  \begin{array}{c} 
  \mathsf{SWA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{S}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{S}
  \end{array} 
  \)
};
\draw[->] (NFA)--(WA);
\draw[->] (NFA)--(SA);
\draw[->] (WA)--(SWA);
\draw[->] (SA)--(SWA);
%\begin{array}{c} \mathsf{NFA} : \Sigma^* \to \mathbb{B} \end{array} 
\end{tikzpicture}
\caption{Classes of Symbolic/Weighted Automata. 
$\Sigma_\mathsf{fin}$ is a finite alphabet, 
$\Sigma_\mathsf{inf}$ is a countable alphabet,
$\mathbb{B}$ is the Boolean algebra, 
$\mathbb{S}$ is an arbitrary commutative semiring, 
$q \xrightarrow{\dots} q'$ represents the form of a transition between states $q$ and $q'$.}
\label{fig:hierarchy}  
\end{figure}

@@

In Section~\ref{section:SWA} we introduce \SWA and \SWT.
Then \SWVPA are defined in Section~\ref{sec:SWVPA}, 
where a polynomial 1-best algorithm is described that can be use to solve
Symbolic Weighted Parsing, as described Section~\ref{sec:parsing}.
Finally, in Section~\ref{sec:transcription}, we present an application 
of this approach to automated music transcription that has been implemented.

@@

\subsection{Definitions} \label{sec:SWTdef}\label{sec:SWAdef}
%(SWT)
\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over the input and output alphabets~$\Sigma$ and $\Delta$ 
and the semiring $\Semiring$ is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\bar{\wei}$ is a tuplet made of the 4 of transition functions 
$\wei_\epsilon$, $\wei_\Sigma$, $\wei_\Delta$, and $\wei_{\Sigma, \Delta}$
from $Q \times Q$ into respectively~$\Semiring$, %$\Phi_\epsilon$, 
$\Phi_\Sigma$, $\Phi_\Delta$, and $\Phi_{\Sigma, \Delta}$.
\end{definition}
%
We synthesize the above 4-uplet of transition functions into a unique 
function~$\wei$ from
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q$
into $\Semiring$, %also called $\wei$ for simplicity, 
such that, for all $q, q' \in Q$, $a \in \Sigma$, and $b \in \Delta$, 
%and with 
%$\< \phi_\epsilon, \phi_\Sigma, \phi_\Delta, \phi_{\Sigma, \Delta}> = \wei(q, q')$, 
\[
\begin{array}{rcll}
\wei(q, \epsilon, \epsilon, q') & = &  \wei_\epsilon(q, q'),\\ %\phi_\epsilon\\
\wei(q, a, \epsilon, q') & = & \phi_\Sigma(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi_\Sigma = \wei_\Sigma(q, q'),\\
\wei(q, \epsilon, b, q') & = & \phi_\Delta(b) & 
\quad\mathrm{where~} \phi_\Delta = \wei_\Delta(q, q'),\\
\wei(q, a, b, q') & = & \phi_{\Sigma, \Delta}(a, b) & 
\quad\mathrm{where~} \phi_{\Sigma, \Delta} = \wei_{\Sigma, \Delta}(q, q').\\
\end{array}      
\]
%More precisely, $Q \times Q$, 
%resp. $Q \times \Sigma \times Q$,
%$Q \times \Delta \times Q$,
%$Q \times \Sigma \times \Delta \times Q$,
%into resp. $\Phi_0$, $\Phi_\Sigma$, $\Phi_\Delta$ $\Phi_{\Sigma, \Delta}$.

The symbolic-weighted transducer $T$ defines a mapping 
from the pairs of strings of $\Sigma^* \times \Delta^*$ 
into the weights of~$\Semiring$,
based on the following intermediate function $\weight_T$
defined recursively for every $q, q' \in Q$, 
%$a \in \Sigma$, $b \in \Delta$ 
for every strings of $s \in \Sigma^*$, $t \in \Delta^*$:
\[
\begin{array}{rccl}
xc(q, s, t, q') & = & 
   \displaystyle\bigoplus_{q'' \in Q} &
   \wei(q, \epsilon, \epsilon, q'') \otimes \weight_T(q'', s, t, q')\\
   & \oplus & \displaystyle\bigoplus_{q'' \in Q} &
              \wei(q'', \epsilon, \epsilon, q') \otimes \weight_T(q, s, t, q'')\\
 & \oplus & \displaystyle\bigoplus_{\begin{array}{c}
                                      \scriptstyle q'' \in Q\\[-2pt]
                                      \scriptstyle s = au, a \in \Sigma
                                      \end{array}} &
   \wei(q, a, \epsilon, q'') \otimes \weight_T(q'', u, t, q')\\
 & \oplus & \displaystyle\bigoplus_{\begin{array}{c}
                                    \scriptstyle q'' \in Q\\[-2pt]
                                    \scriptstyle t = bv, b \in \Delta
                                    \end{array}} &
    \wei(q, \epsilon, b, q'') \otimes \weight_T(q'', s, v, q')\\ 
 & \oplus & \displaystyle\bigoplus_{\begin{array}{c}
                                    \scriptstyle q'' \in Q\\[-2pt]
                                    \scriptstyle s = au, a \in \Sigma\\[-2pt]
                                    \scriptstyle t = bv, b \in \Delta
                                    \end{array}} &
   \wei(q, a, b, q'') \otimes \weight_T(q'', u, v, q').\\ 
\end{array}
\]
The sum above can be infinite because of the expression in the first line, 
however, $\T(\Omega)$ is well defined since $\Semiring$ is complete.
%
As we shall see, epsilon-transitions can be removed and hence the sum made finite.
%
%** The application of $\oplus$ can be seen as a ND choice between transitions **
We recall that, by convention, an empty sum with $\oplus$ is $\zero$. 
%
Since $\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei(q, a, b, q'')$ equal to $\zero$ in the above expression 
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and writing $b$ 
(including the case $a = \epsilon$ and $a = \epsilon$).
%These functions $\phi$ act as guards for the transducer's transitions, 
%preventing a transition when they return the absorbing $\zero$ of $\Semiring$.
This is analogous to a transition's guard not satisfied by $\<a, b>$ in 
the case of symbolic transducers.

@@

\noindent
The \emph{Symbolic Weighted Automata} %$A = \< Q, \init, \weight, \final >$
over $\Sigma$ and $\Semiring$ 
are defined similarly as the transducers of Definition~\ref{def:SWT}, 
by simply omitting the output symbols.
%
In this case, a label theory over $\Sigma$ is reduced to
a set $\Phi_\Sigma$ closed under~$\oplus$ and~$\otimes$.
%
\begin{definition} \label{def:SWA}
A \emph{symbolic-weighted automaton} (\SWA)
over the input alphabet~$\Sigma$
and the commutative semiring $\Semiring$ is a tuple
$A = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\bar{\wei}$ is a pair of transition functions 
$\wei_\epsilon$ and $\wei_\Sigma$ from $Q \times Q$ into 
respectively $\Semiring$ %$\Phi_\epsilon$ 
and $\Phi_\Sigma$.
\end{definition}

@@

\begin{proposition}
Given a \SWT $T$ 
over alphabet $\Sigma$, $\Delta$ and a commutative semiring~$\Semiring$, 
and $s \in \Sigma^*$, 
there exists an effectively constructible \SWA 
$A_{T, s}$ over $\Delta$ and $\Semiring$ 
such that for all $t \in \Delta^*$, $A_{T, s}(t) = T(s, t)$.
\end{proposition}
%
\begin{proof}
Let $T = \< Q, \init, \bar{\wei}, \final >$ 
where $\bar{\wei} \< \wei_\epsilon, \wei_\Sigma, \wei_\Delta, \wei_{\Sigma, \Delta}>$
has an associated synthesized form $\wei$, 
and let $s = s_1 \ldots s_n$ with $s_1, \ldots, s_n \in \Sigma$.
%$n = |s|$ be the length of $s$.
%
Let $Q' = [1..n] \times Q$ be the state set of $A_{T, s}$,
and let, for all $q \in Q$, 
$\init'(1, q) = \init(q)$ and $\init'(i, q) = \zero$ for all $1 < i \leq n$, 
and $\final'(n, q) = \final(q)$ and $\final'(i, q) = \zero$ for all $1 \leq i <  n$.

\noindent
We define the functions $\wei'_\epsilon$ and $\wei'_\Sigma$ for $A_{T, s}$
in their synthesized form: for all $1 \leq i, j \leq n$, $q, q' \in Q$ and $a \in \Sigma$, 
and $b \in \Delta$, 
\[
\begin{array}{lclcl}
\wei'(\< i, q>, \epsilon, \< i, q'>)   & = & \wei(q, \epsilon, \epsilon, q'), & \quad & \\ 
\wei'(\< i, q>, \epsilon, \< i+1, q'>) & = & \wei(q, s_i, \epsilon, q') & & \mathrm{if~} i < n,\\ 
\wei'(\< i, q>, \epsilon, \< j, q'>)   & = & \zero & & \mathrm{if~} j \neq i, j \neq i+1,\\ 
\wei'(\< i, q>, b, \< i, q'>)          & = & \wei(q, \epsilon, b, q'),\\ 
\wei'(\< i, q>, b, \< i+1, q'>)        & = & \wei(q, s_i, b, q') & & \mathrm{if~} i < n,\\ 
\wei'(\< i, q>, b, \< j, q'>)          & = & \zero & & \mathrm{if~} j \neq i, j \neq i+1. 
\end{array}      
\]
Note that the functions defined above belong to the set $\Phi_\Delta$ 
of the underlying label theory, 
by the hypothesis of closure under partial application.
%
The $\SWA$ wanted is $A_{T, s} = \< Q', \init', \bar{\wei'}, \final' >$.
\qed
\end{proof}

@@

\subsection{Definition}
Let $\Omega$ be a countable alphabet 
%finite (large) or infinite,
that we assume partitioned into three subsets 
$\Omegai$, $\Omegac$, $\Omegac$,
whose elements are respectively called 
\emph{internal}, \emph{call} and \emph{return} symbols.
% \begin{itemize}
% \item a set $\Omegai$ of \emph{internal symbols} denoted $a$,
% \item a set $\Omegac$ of \emph{call symbols} denoted $\call{a}$,
% \item a set $\Omegar$ of \emph{return symbols} denoted $\return{a}$.
% \end{itemize}
Let~$\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative semiring and  
let  $(\Phi_\epsilon, \Phic, \Phir, \Phicr)$ be a label theory over $\Semiring$
%In order to simplify notations, %and following the definition of Section~\ref{section:transducer}, 
%we shall write respectively 
where $\Phic$, $\Phir$ and~$\Phicr$ stand respectively 
for~$\Phi_\Omegac$, $\Phi_\Omegar$ and~$\Phi_{\Omegac, \Omegar}$.
%
Moreover, we extend this theory with a set $\Phii$ 
of unary functions in $\Omegai \to \Semiring$,
closed under $\oplus$ and $\otimes$.

@@

\noindent
The intuition is the following for the above transitions.
\begin{description}
\item $\weii$ : read the input internal symbol $a$, change state to $q'$ (stack is untouched).
\item $\weic$ : read the input call symbol $\call{c}$, push it to the stack along with $p$, change state to $q'$.
\item $\weir$ : when the stack is not empty, 
      read and pop from stack a pair made of $\call{c}$ and $p$, 
      read the input return symbol $\return{r}$, change state to $q'$.
      In this case, the weight function $\phi_\mathsf{r}$ 
      computes a value of matching between the call and return symbols $c$ and $r$.
      This value might be set to $\zero$ in order to express that the symbols do not match.
\item $\weie$ : when the stack is empty, 
      read the input symbol $\call{r}$, change state to $q'$.
\end{description}

We give now a formal definition of these transitions of the automaton~$A$
in term of a weight value
computed by an intermediate function $\weight_A$, like in Section~\ref{sec:SWT}.
In the case of a pushdown automaton, a configuration is composed 
of a state $q \in Q$ and a stack content $\gamma \in \Gamma^*$, where $\Gamma = \Omegac \times P$.
Therefore, $\weight_A$ is a function from 
$Q \times \Gamma^* \times \Omega^* \times Q \times \Gamma^*$ into $\Semiring$.
\[
\begin{array}{rcl}
\weight_A\bigl(\config{q}{\gamma}, a\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weii(q, a, q'') 
 \otimes \weight_A\bigl(\config{q''}{\gamma}, u, \config{q'}{\gamma'}\bigr)\\
%
\weight_A\bigl(\config{q}{\gamma}, {\call{c}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-2pt]
                          \scriptstyle p \in P
                          \end{array}}}
 \weic\bigl(q, {\call{c}}, q'', p\bigr) 
 \otimes \weight_A\bigl(\config{q''}{{\call{c}}\, p\cdot \gamma}, u, \config{q'}{\gamma'}\bigr)\\[1mm]
%
\weight_A\bigl(\config{q}{{\call{c}}\, p\cdot \gamma}, {\return{r}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} 
 \weir\bigl(q, {\call{c}}, p, {\return{r}}, q''\bigr) 
 \otimes \weight_A\bigl(\config{q''}{\gamma}, u, \config{q'}{\gamma'}\bigr)\\
%
\weight_A\bigl(\config{q}{\bot}, {\return{r}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weie(q, {\return{r}}, q'') 
 \otimes \weight_A\bigl(\config{q''}{\bot}, u, \config{q'}{\gamma'}\bigr)\\
\end{array}
\]
where $\bot$ denotes the empty stack and ${\call{c}}\, p\cdot \gamma$ 
denotes a stack where the pair made of ${\call{c}} \in \Omegac$ and $p \in P$ is the top symbol 
and $\gamma$ is the rest of stack.

@@

\section{Trees and Nested-Words}
see also \cite{Caralp12VPAmult} for a discussion on the correspondence 
between tree automata and VPA.
